Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-09-07 ms.tex
Updates to the DM section

\altaffiltext{1}{Corresponding authors: N. Karlsson, nkarlsson@physics.umn.edu \& C. Pfrommer, christoph.pfrommer@h-its.org}
\altaffiltext{2}{School of Physics and Astronomy, University of Minnesota, Minneapolis, MN 55455, USA}
\altaffiltext{3}{DESY, Platanenallee 6, 15738 Zeuthen, Germany}
\altaffiltext{4}{Institut f\"ur Physik und Astronomie, Universit\"at Potsdam, 14476 Potsdam-Golm, Germany}
\altaffiltext{5}{Heidelberg Institute	for Theoretical	Studies, Schloss-Wolfsbrunnenweg 35, D-69118 Heidelberg, Germany}
\altaffiltext{67}{Department of Physics, University of California, Santa Barbara, CA 93106, USA}

@@

\begin{abstract}
Theoretical models of galaxy clusters suggest they may be sources of very high-energy (VHE) gamma-ray emission. Observations of radio halos and relics in clusters indicate populations of relativistic electrons, which are either injected into the intra-cluster medium and accelerated by turbulent shocks or secondaries from interactions of cosmic-ray hadrons (protons and ions) with the intra-cluster gas. Those interactions would also produce gamma rays through the decay of neutral pions. Reported here are observations of the Coma cluster of galaxies with the VERITAS array of imaging Cherenkov telescopes. The observations were made between March and May 2008. Radio observations at 1.41 GHz taken with the Green Bank Telescope and X-ray observations with the ROSAT All-Sky survey was used to a priori select regions of interest in the field of view. No significant VHE gamma-ray emission from the Coma cluster was detected by VERITAS. Several emission regions are considered, including the cluster core as both unresolved (point-like) and spatially extended, and a select few cluster members. The flux upper limits at the 99\% confidence level measured are on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$ s$^{-1}$, which is two orders of magnitude higher than model-predicted minimum gamma-ray fluxes.

The gamma-ray flux upper limits have been used to constrain the populations of cosmic-ray particles and magnetic fields and in the Coma cluster of galaxies. The relative cosmic-ray pressure, $X_{\mathrm{CR}}=\expval{P_{\CR}}/\expval{P_{\rmn{th}}}$, is \ldots. The observations imply a lower limit on the intra-cluster magnetic field on the order of a $\mu$G, depending on the magnetic decline and gamma-ray spectral index. Finally, a constraint on the thermally averaged product of the total self-annihilation cross section and the relative velocity of the dark-matter particles, $\expval{\sigma v}$, is derived and found to be $\lesssim 10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}\gtrsim 1000$ GeV/c$^{2}$.
\end{abstract}

@@

\item In the last step, we solve for $B_{0}$ subject to the boundary condition to match the observed synchrotron profile while fixing the profile of $X_{\CR}(r)$ as determined through the previous two steps. Note that for $B_{0} \gg B_{\rmn{CMB}}$ and a radio spectral index of $\alpha_{\nu}=1$, the solution would be degenerate.
\end{enumerate}

@@

\section{Emission from Dark Matter Annihilations}
When the size of the integration region is increased, the limits on $\expval{\sigma v}$ result from a competition between the gain in the astrophysical factor $\expval{J}$ and the integrated background. For integration regions larger than 0.2$^{\circ}$ in radius, the astrophysical factor no longer compensate for the increased count of background events and the signal-to-noise deteriorates.

\section{Discussion and Conclusions}
VERITAS observed the Coma cluster of galaxies for a total of 18.6 hours of high-quality live time between March and May in 2008.

We employed various approaches to constrain the CR population and magnetic field distribution that are complementary in their assumptions and hence well suited in assessing underlying Bayesian priors in the models. (1) We use a simplified ``isobaric CR model'' that is characterized by a constant CR-to-thermal pressure fraction and has a power-law momentum spectrum.  While this model is physically not justified a priory, it is simple and widely used in the literature and captures some aspects of more elaborate models such as (2) the simulation-based analytical approach of \citet{article:PinzkePfrommer:2010}. This is a ``first principle approach'' that predicts the CR distribution for a given set of assumptions (namely the physics that was simulated). It is powerful since it only requires the density profile as input due to the approximate universality of the CR distribution (when neglecting CR diffusion and streaming which appears, however, necessary to explain the radio halo bimodality). (3) We finally use a pragmatic approach which models the CR and magnetic distribution such as to reproduce the observed emission profile of the Coma radio halo. While this approach is also physically not justified, it is powerful as it shows what the ``correct'' model has to achieve and can point into the direction of the relevant physics.

>>>>

\altaffiltext{1}{Corresponding authors: N. Karlsson, nkarlsson@physics.umn.edu \& C. Pfrommer, christoph.pfrommer@h-its.org}
\altaffiltext{2}{School of Physics and Astronomy, University of Minnesota, Minneapolis, MN 55455, USA}
\altaffiltext{3}{DESY, Platanenallee 6, 15738 Zeuthen, Germany}
\altaffiltext{4}{Institut f\"ur Physik und Astronomie, Universit\"at Potsdam, 14476 Potsdam-Golm, Germany}
\altaffiltext{5}{Heidelberg Institute	for Theoretical	Studies, Schloss-Wolfsbrunnenweg 35, D-69118 Heidelberg, Germany}
\altaffiltext{6}{Department of Physics, University of California, Santa Barbara, CA 93106, USA}

@@

\begin{abstract}
Theoretical models of galaxy clusters suggest they may be sources of very high-energy (VHE) gamma-ray emission. Observations of radio halos and relics in clusters indicate populations of relativistic electrons, which are either injected into the intra-cluster medium and accelerated by turbulent shocks or secondaries from interactions of cosmic-ray hadrons (protons and ions) with the intra-cluster gas. Those interactions could also produce gamma rays through the decay of neutral pions. Reported here are observations of the Coma cluster of galaxies with the VERITAS array of imaging Cherenkov telescopes. The observations were made between March and May 2008. Radio observations at 1.41 GHz taken with the Green Bank Telescope and X-ray observations with the ROSAT All-Sky survey was used to a priori select regions of interest in the field of view. No significant VHE gamma-ray emission from the Coma cluster was detected by VERITAS. Several emission regions are considered, including the cluster core as both unresolved (point-like) and spatially extended, and a select few cluster members. The flux upper limits at the 99\% confidence level measured are on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$ s$^{-1}$, which is two orders of magnitude higher than model-predicted minimum gamma-ray fluxes.

The gamma-ray flux upper limits have been used to constrain the populations of cosmic-ray particles and magnetic fields and in the Coma cluster of galaxies. The relative cosmic-ray pressure, $X_{\mathrm{CR}}=\expval{P_{\CR}}/\expval{P_{\rmn{th}}}$, is on the order of 10-20\% when assuming a gamma-ray photon index $\alpha=2.3$. The observations imply a lower limit on the intra-cluster magnetic field on the order of a $\mu$G, depending on the magnetic decline and gamma-ray spectral index. Finally, a constraint on the thermally averaged product of the total self-annihilation cross section and the relative velocity of the dark-matter particles, $\expval{\sigma v}$, is derived and found to be $\lesssim 10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}\gtrsim 1000$ GeV/c$^{2}$.
\end{abstract}

@@

\item In the last step, we scale the secondary radio emission profile linearly with $X_{\CR,0}$ and solve for $B_{0}$ subject to the boundary condition to match the observed synchrotron profile. Note that for $B_{0} \gg B_{\rmn{CMB}}$ and a radio spectral index of $\alpha_{\nu}=1$, the solution would be degenerate.
\end{enumerate}

@@

\section{Emission from Dark Matter Annihilations}
As already mentioned in the introduction, most of the mass in a galaxy cluster is in the form of DM. While the nature of this DM remains unknown, the most compelling theoretical candidate is a WIMP. The self-annihilations of WIMPs are expected to produce a flux of secondary gamma rays with a spectral signature that deviates significantly from the power-law spectrum observed in most conventional astrophysical sources. The spectrum will also feature a very sharp cutoff at the WIMP mass. Observations of these spectral features together with the expected difference in spatial distribution of the gamma-ray flux compared to conventional astrophysical sources allows for a unique means to indirectly detect DM.

The gamma-ray flux due to self-annihilations of WIMPs is given by 
\begin{equation}
\frac{dN(\Delta\Omega)}{dE}=\frac{\expval{\sigma v}}{8\pi m_{\chi}^{2}}\left[\frac{dN(E,m_{\chi})}{dE}\right]\int_{\Delta\Omega}d\Omega\int d\lambda \rho^{2}(\lambda,\Omega),
\label{eqn:WIMPflux}
\end{equation}
where $\expval{\sigma v}$ is the thermally averaged product of the total self-annihilation cross section and the WIMP velocity, $m_{\chi}$ is the WIMP mass, $dN(E,m_{\chi})/dE$ is the gamma-ray yield per annihilation, $\Delta\Omega$ is the observed solid angle, $\rho$ is the DM mass density, and $\lambda$ is the line-of-sight distance to the integration volume.

The two integrals in Eq. (\ref{eqn:WIMPflux}) results in a dimensionless factor representing the astrophysical contribution to the flux
\begin{equation}
J(\Delta\Omega)=\left(\frac{1}{\rho_{c}^{2}R_{H}}\right) \int_{\Delta\Omega}d\Omega\int d\lambda \rho^{2}(\lambda,\Omega),
\end{equation}
which here has been normalized to the square of the critical density, $\rho_{c}$, and the Hubble radius, $R_{H}$.

Given the upper limit on the observed gamma-ray rate, $R_{\gamma}(95\%\mathrm{CL})$, we can place constraints on the WIMP parameter space $(m_{\chi}, \expval{\sigma v})$.

%When the size of the integration region is increased, the limits on $\expval{\sigma v}$ result from a competition between the gain in the astrophysical factor $\expval{J}$ and the integrated background. For integration regions larger than 0.2$^{\circ}$ in radius, the astrophysical factor no longer compensate for the increased count of background events and the signal-to-noise deteriorates.

\section{Discussion and Conclusions}
VERITAS observed the Coma cluster of galaxies for a total of 18.6 hours of high-quality live time between March and May in 2008. No significant excess of gamma-rays above a threshold of about 220 GeV was detected. Flux upper limits at 99\% confidence level were calculated for the cluster core and three member galaxies. The core was considered both as a point-like and spatialy extended source. The obtained flux upper limits were used to constrain properties of the cluster.

We employed various approaches to constrain the CR population and magnetic field distribution that are complementary in their assumptions and hence well suited in assessing underlying Bayesian priors in the models. (1) We use a simplified ``isobaric CR model'' that is characterized by a constant CR-to-thermal pressure fraction and has a power-law momentum spectrum.  While this model is physically not justified a priori, it is simple and widely used in the literature and captures some aspects of more elaborate models such as (2) the simulation-based analytical approach of \citet{article:PinzkePfrommer:2010}. This is a ``first principle approach'' that predicts the CR distribution for a given set of assumptions (namely the physics that was simulated). It is powerful since it only requires the density profile as input due to the approximate universality of the CR distribution (when neglecting CR diffusion and streaming which appears, however, necessary to explain the radio halo bimodality). (3) We finally use a pragmatic approach which models the CR and magnetic distribution such as to reproduce the observed emission profile of the Coma radio halo. While this approach is also physically not justified, it is powerful as it shows what the ``correct'' model has to achieve and can point into the direction of the relevant physics.
